---
title: "Modeling"
author: "Group 6"
output: pdf_document
---

```{r setup, include=FALSE}
# setup chunks
knitr::opts_chunk$set(echo = F, fig.align = "center", warning = F, message = F, fig.height = 4)
# load libaries
library(tidyverse)
library(grid)
library(gridExtra)
library(patchwork)
library(ISLR)
# load helper scripts
source(file = "../R/wrangle.R")
source(file = "../R/model.R")
source(file = "../R/visualization.R")
source(file = "../R/threshold.R")
# global parameters
bplot <- T
bloud <- T
bhead <- T
```

```{r}
#=============================#
#       Import Datasets       #
#=============================#
# Load rookies datasets
df_pit_rkes <- read_csv("../data/rookie-pitcher.csv")
df_pos_rkes <- read_csv("../data/rookie-position.csv")
# Load retirees datasets
df_pit_ret <- read_csv("../data/retirees-pitcher.csv")
df_pos_ret <- read_csv("../data/retirees-position.csv")
# Find number of retirees by year
num_retirees <- total_retirees_by_yr(df_pit_ret, df_pos_ret)
num_retirees <- data.frame(retirees = num_retirees$retirees)
```

```{r, cache = T}
#==============================#
#       Position Players       #
#==============================#
# Obtain wrangled datasets
pos_rkes <- wrangle_init(df_pos_rkes)
pos_ret <- wrangle_init(df_pos_ret)
# Get thresholds in each year
pos_rkes_thresholds <- find_thresholds(pos_rkes)
# See and record which players cross that year's threshold
pos_ret <- compare_thresholds(pos_ret, pos_rkes_thresholds)
```

```{r, cache = T}
#==============================#
#        Pitcher Players       #
#==============================#
# Obtain wrangled datasets
pit_rkes <- wrangle_init(df_pit_rkes)
pit_ret <- wrangle_init(df_pit_ret)
# Get thresholds in each year
pit_rkes_thresholds <- find_thresholds(pit_rkes)
# See and record which players cross that year's threshold
pit_ret <- compare_thresholds(pit_ret, pit_rkes_thresholds)
```

```{r}
write.csv(pit_ret,"../data-gen/pit_ret_classified.csv", row.names = F)
write.csv(pos_ret,"../data-gen/pos_ret_classified.csv", row.names = F)
```

# The Couldabeen Classification Problem

```{r, fig.height = 6}
# Density plot of rookie and retired pitchers 
pitcher_density <- ggplot() +
  geom_density(data = pit_ret, aes_string(x = "WAR"), color = "salmon") +
  geom_density(data = pit_rkes, aes_string(x = "WAR"), color = "skyblue3") +
  geom_vline(xintercept = median(pit_rkes$WAR), color = "skyblue3") +
  labs(title = "Pitchers")
# Density plot of rookie position players and retired position players 
position_density <- ggplot() +
  geom_density(data = pos_ret, aes_string(x = "WAR"), color = "salmon") +
  geom_density(data = pos_rkes, aes_string(x = "WAR"), color = "skyblue3") +
  geom_vline(xintercept = median(pos_rkes$WAR), color = "skyblue3") +
  labs(title = "Position")
# Plot
if(bplot){grid.arrange(pitcher_density, position_density, nrow = 2)}
```

\newpage

# Counting Couldabeens

```{r, echo = T}
#===================================#
#        Counting: Couldabeens      #
#===================================#
# Combine the threshold-classified retiree datasets
retirees <- rbind(pit_ret,pos_ret)
# Count couldabeens
couldabeens <- count_cbns(retirees)
```

Our retirees dataframe looks like this:

```{r, echo = F}
head(retirees)
```

Our couldabeens dataframe looks like this:

```{r, echo = F}
if(bhead){head(couldabeens)}
```

## First Look: A Logistic Model

```{r}
# Partition retirees into pre-rule and post-rule era
retirees_pre <- prerule(retirees)
retirees_post <- postrule(retirees)
```

```{r, fig.width = 6, fig.height = 4}
# On the retirees pre-rule dataset
if(bplot){plot_logmodel(retirees_pre, "Retirees Above and Below Threshold (Pre-rule)")}
model_log <- logistic_model(retirees_pre)
if(bloud){summary(model_log)}
```

\newpage

```{r, fig.width = 6, fig.height = 4}
# On the retirees pre-rule dataset
dataset <- retirees_post
if(bplot){plot_logmodel(dataset, "Retirees Above and Below Threshold (Post-rule)")}
model_log <- logistic_model(dataset)
if(bloud){summary(model_log)}
```

\newpage

# Computing Retiree Proportions

```{r, echo = T}
#===================================#
#      Proportions: Couldabeens     #
#===================================#
# Find number of retirees by year
num_retirees <- total_retirees_by_yr(df_pit_ret, df_pos_ret)
num_retirees <- data.frame(retirees = num_retirees$retirees)
# Append number of retirees that year
couldabeens <- cbind(couldabeens, num_retirees)
# Find proportion of couldabeens : retirees
couldabeens <- couldabeens %>% mutate(prop = cbns/retirees)
```

Here is what the proportion-appended couldabeen dataframe looks like:

```{r, echo = F}
if(bhead){head(couldabeens)}
```

\newpage

# Year as Predictor: Linear Modeling

```{r, echo = T}
#==================#
#     Modeling     #
#==================#
# Partition dataset into years before and after rule
couldabeens_pre <- prerule(couldabeens)
couldabeens_post <- postrule(couldabeens)
# Obtain linear model for pre-rule years
model_pre <- linear_model(couldabeens_pre)
coefs_pre <- model_pre$coefficients
# Obtain linear model for post-rule years
model_post <- linear_model(couldabeens_post)
coefs_post <- model_post$coefficients
# Obtain linear model for all years
model_comp <- linear_model(couldabeens)
coefs_comp <- model_comp$coefficients
```

\newpage 

## Couldabeens: A Comprehensive Look

```{r}
## + geom_abline(slope = coefs_comp[2], intercept = coefs_comp[1], color = col) + 
if(bloud){summary(model_comp)}
```

```{r}
col <- "violet"
col1 <- "salmon"
col2 <- "skyblue2"
# Plot proportion of couldabeens in post-rule era ()
plot_comp <- scatter_props(couldabeens, title = "Proportion of Couldabeen Retirees") +
    geom_abline(slope = coefs_comp[2], intercept = coefs_comp[1], color = col) + 
    geom_point(data = couldabeens_pre, mapping = aes(x = Year, y = prop), color = col1) +
    geom_abline(slope = coefs_pre[2], intercept = coefs_pre[1], color = col1) +
    geom_point(data = couldabeens_post, mapping = aes(x = Year, y = prop),color = col2) +
    geom_abline(slope = coefs_post[2], intercept = coefs_post[1], color = col2) + 
    labs(y = "Proportion")
plot_comp
```


\newpage

## Couldabeens: Pre-rule Era (1969-2002)

```{r}
if(bloud){summary(model_pre)}
```

\bigskip
\bigskip

```{r, fig.cap= "Proportion of Retirees who were Coulabeens prior to the implementation of the Luxury Tax"}
# Plot proportion of couldabeens in pre-rule era ()
plot_pre <- plot_props_linear(couldabeens_pre, coefs_pre, color = "salmon", 
                              title = "Proportion of Couldabeen Retirees (1969-2002)")
plot_pre
```

\newpage

## Couldabeens: Post-rule Era (2003-2018)

```{r}
if(bloud){summary(model_post)}
```

\bigskip
\bigskip

```{r, fig.cap= "Proportion of Retirees who were Coulabeens after the implementation of the Luxury Tax"}
# Plot proportion of couldabeens in post-rule era ()
plot_post <- plot_props_linear(couldabeens_post, coefs_post, color = "skyblue2", 
                               title = "Proportion of Couldabeen Retirees (2003-2018)")
plot_post
```

\newpage

# Quadratic Regression Model

```{r}

```


